---
title: "Reproducible Research: Peer Assessment 1"
author: "Melanie Hardie"
output: 
  html_document:
    keep_md: true
---
In this assignment, we are going to explore data collected from a personal activity monitoring device for a two month period. But before we begin, we need to load the data from a CSV file.

## Loading and preprocessing the data

We assume that a zip file named `"activity.zip"` is located in this directory. However, if it is not, we will download it from the link provided by our course instructors. Note that we are running RStudio Version 0.99.467 on OSX Version 10.10.4.  

```{r}
# if the zip file cannot be found in this directory, go get it
if (!file.exists("activity.zip")) {
    fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(fileUrl, destfile = "activity.zip", method = "curl")
}
# unzip - we will now have a file called "activity.csv"
unzip("activity.zip")
# read in the data
activityData <- read.csv("activity.csv", stringsAsFactors = FALSE)
# convert the data variable to the Date class
activityData$date <- as.Date(activityData$date)
# summary of data
str(activityData)
```
  
  From the output above, we see that `activityData` is a data.frame of 17568 observations. The 3 variables are `'steps'`, `'date'`, and `'interval'`, which are of the classes `int`, `Date`, and `int`, respectively. Now let's play!  

## What is mean total number of steps taken per day?

First, we are interested in seeing the total number of steps taken per day.

```{r}
# sum the number of steps for each day
stepsByDay <- aggregate(activityData$steps ~ activityData$date, FUN=sum, na.rm=TRUE)
# rename the columns
colnames(stepsByDay) <- c("day", "totalSteps")
# create a histogram and add the x-axis ticks and labels
hist(stepsByDay$totalSteps, col = "green", breaks = 20, main = "Histogram of Total Steps Taken per Day", 
     xaxt='n', xlim = c(0,22000), xlab = "Total Steps Taken per Day")
axis(side=1, at=seq(0,22000,2000), labels=seq(0,22000,2000))
```

The code above created a data frame called `stepsByDay` that has columns `day` and `totalSteps`. We can use this data structure to find the mean and median number of steps taken each day:

```{r}
mean(stepsByDay$totalSteps)
median(stepsByDay$totalSteps)
```
We could have also obtained these numbers using the `summary` function. Let's see how the results compare:

```{r}
# print an appropriate number of digits
options(digits=10)
summary(stepsByDay$totalSteps)
```

We have confirmed that the mean total number of steps taken per day is 10766.19 and the median total number of steps taken per day is 10765.  

## What is the average daily activity pattern?

Now we would like to explore possible daily activity patterns and to do that we will consider the `interval` variable. We are told that this variable is coded as the 5 minute interval in which the measurement was taken. Let's see what the values look like:

```{r}
range(activityData$interval)
unique(activityData$interval)
```

We observe that the values range from 0 to 2355. The values start a 0, then increment by 5 until 55, then the next value is 100. This same pattern repeats within each hundreds interval. For example, after 100, the values increment by 5 until 155, and then jump to 200. We will assume this is coded as the interval start time, described by the 24 hour clock, which can be padding with leading 0's to form a four-digit number that represents the hour and minute. For example, 105 would be padded with one leading 0 to form 0105. The first two numbers represent the hour, so 01 (or 1 am), and the last two numbers represent the minute 05. So the interval value of 105 is the observation corresponding to the 5 minute interval starting at 1:05 am.  

We would like to see if there are any daily trends in these intervals, so we create a time series plot of each 5 minute interval and the average number of steps taken during that interval across all the days.

```{r}
# average the number of steps for each interval
stepsByInterval <- aggregate(activityData$steps ~ activityData$interval, FUN=mean, na.rm=TRUE)
# rename the columns
colnames(stepsByInterval) <- c("interval", "averageSteps")
# create a line plot of the average steps taken per interval
library(ggplot2)
ggplot(stepsByInterval, aes(interval, averageSteps)) + geom_line(col="steelblue",lwd=1.5) + theme_bw() + 
    coord_cartesian(xlim = c(0,2355)) + scale_x_continuous(breaks=seq(0, 2400, 200)) + 
    labs(x="Interval",y="Average number of steps taken", title="Average Number of Steps Taken per Interval")
```

From the plot above we can see a distinct spike in number of steps taken between approximately 8 and 9 am. We can find the observation with the highest average number of step taken using the `max()` function:

```{r}
subset(stepsByInterval, stepsByInterval$averageSteps==max(stepsByInterval$averageSteps))
```
Therefore, on avergae, interval 835 corresponding to the 5-minute interval starting at 8:35 am contains the maximum number of steps.

## Imputing missing values

We noticed at the onset of our exploration from the output of the `str()` function that there we several NA values reported for the `steps` variable. We can count the number of NA's in `steps` using:

```{r}
sum(is.na(activityData$steps))
```

2304 NA's are reported in the `steps` column. If we do the same for the other two variables, we can confirm that there are no NAs in either of the other two columns.

```{r}
sum(is.na(activityData$interval))
sum(is.na(activityData$date))
```

We coud have also made this conclusion using `summary`, as shown below:
```{r}
summary(activityData)
```

Since we see on the bottom line of the output that there are 2304 NA's reported in the data set and we have determined that there are 2304 NA's in the `steps` column, we know the other columns must be complete.

We would like to replace the NA's with some value to see how this will affect the total number of steps taken each day. Since we have seen that the number of steps taken seems to vary according to the time of day, we will fill in the missing values using the values computed above - the average number of steps for each 5-minute interval. 

```{r}
# create a copy of activityData 
imputedActivityData <- activityData
# add the average interval steps as a column to the data frame
imputedActivityData <- cbind(imputedActivityData,stepsByInterval$averageSteps)
# find the rows with NA's
naIndexVector <- which(is.na(imputedActivityData))
# replace the NA with average steps for that 5-minute interval
imputedActivityData[naIndexVector,]$steps <- imputedActivityData[naIndexVector,4]
# we should now find that there are 0 NAs in the steps column
sum(is.na(imputedActivityData$steps))
```

Success!

Let's take another look at a histogram of the total steps taken per day now that we have replaced the missing values:

```{r}
# sum the number of steps for each day
stepsByDay <- aggregate(imputedActivityData$steps ~ imputedActivityData$date, FUN=sum, na.rm=TRUE)
# rename the columns
colnames(stepsByDay) <- c("day", "totalSteps")
# create a histogram and add the x-axis ticks and labels
hist(stepsByDay$totalSteps, col = "cyan", breaks = 20, 
     main = "Histogram of Total Steps Taken per Day \nusing Imputed Values", xaxt='n', 
     xlim = c(0,22000), xlab = "Total Steps Taken per Day")
axis(side=1, at=seq(0,22000,2000), labels=seq(0,22000,2000))
```

Again, we can use the data frame we created in the code above called `stepsByDay` to find the mean and median number of steps taken each day:

```{r}
mean(stepsByDay$totalSteps)
median(stepsByDay$totalSteps)
```
Compared to output from the `summary` function:

```{r}
# print an appropriate number of digits
options(digits=10)
summary(stepsByDay$totalSteps)
```
So we find that the mean and median are both now 10766.19 (rounded where applicable). Recall that this is the same mean value we computed before replacing the NA's. The median has incresed by a little more than one step from 10765 to 10766.19. This demonstrates that imputing the missing values has had little impact on the estimates of the total daily number of steps.  

## Are there differences in activity patterns between weekdays and weekends?

Our final analysis with this data set will be to compare the average steps taken on weekdays to weekends. First, we will add a factor variable to the imputed data, then we will create a plot similar to the Average Number of Steps Taken plot above. However, this time we will create separate plots for weekdays and weekends.
```{r}
# create the a vector of weekend days
weekendDays <- c("Saturday","Sunday")
# add a new column to the imputed data set based on
# whether the day is in the weekendDays vector
imputedActivityData$dayFactor <- ifelse(weekdays(imputedActivityData$date) %in% weekendDays,"weekend","weekday")
# convert to factor
imputedActivityData$dayFactor <- as.factor(imputedActivityData$dayFactor)
# average the number of steps for each interval by factor
stepsByInterval <- aggregate(imputedActivityData$steps,
                             by=list(imputedActivityData$interval,imputedActivityData$dayFactor), 
                             FUN=mean, na.rm=TRUE)
# rename the columns
colnames(stepsByInterval) <- c("interval", "Factor", "averageSteps")
# create a line plot of the average steps taken per interval for factors
ggplot(stepsByInterval, aes(interval, averageSteps, colour = Factor)) + geom_line(lwd=1.5) + 
    theme_bw() + coord_cartesian(xlim = c(0,2355)) + facet_grid(Factor ~ .) +
    scale_x_continuous(breaks=seq(0, 2400, 200)) +
    labs(x="Interval",y="Average number of steps taken", title="Average Number of Steps Taken per Interval")
```

From the plots above, we can see that the 8 - 9 am spike in steps is mostly due to weekdays and the average number of steps is more evenly distributed among the intervals on the weekend. It also seems that the test subject is an early bird who get up a little before 6 am - *shudder*!